---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Ethan Putnam"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
moco_scores <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- moco_scores |> summarize(mean(compliance_score), sd(compliance_score), min(compliance_score), max(compliance_score))
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
moco_scores |>
  ggplot() +
  geom_histogram(aes(x = compliance_score), bindwidth = 2) +
   geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Montgomery County Restaurant Health Inspection Compliance Scores",
    x = "Compliance Score",
    y = "Count"  )
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**

The standard deviation of 5.8 paired with the histogram show that the distribution is not even, and is skewed heavily left. The vast majority of restaurants being fully complicit is not the story, but the outliers that were less complicit are newsworthy. The next step would be examining the restaurants that were least complicit.

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |> summarize(mean(bus), sd(bus), mean(rail), sd(rail))
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}

total_bus = wmata_daily |> count(bus)
total_rail = wmata_daily |> count(rail)

wmata_zscores <- wmata_daily |> mutate(rail_zscore = (rail - mean(rail)) / sd(rail), bus_zscore = ((bus - mean(bus)) / sd(bus)))

```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}

unusual_scores <- wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))

```

There are six unusual days.

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(50)

sample_91 <- wmata_daily |> sample_n(91)

sample_91 |> summarize(mean(bus), sd(bus), mean(rail), sd(rail))
```

The means and standard deviations are similar, but all a bit off. This suggests that a sample of a large dataset will provide mostly accurate summary of the dataset, but only to an extent. A larger sample will always yield more accurate results to the larger dataset.

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)


Four of the six days were holidays, which makes teh reduced travel make sense. The other two days both fell on 1/6, so investigating why travel by bus and train was down on January 6th in both 2024 and 2025 could be a story.

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts |> mutate(theft_rate22 = (thefts_2022/population) * 10000, theft_rate23 = (thefts_2023/population) * 10000)
```

Per capita rates are more useful than raw counts because they can be compared between counties of varying sizes much more accurately than raw counts.

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts |> summarize(median(theft_rate22), median(theft_rate23))
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

I would prioritize using a z-score to determine how great the outliers are from the mean, given that a standard deviation of 42 is substantial and suggests that there are some major outliers.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

Similar to the procedure in task 8, pools could be filtered by which have z-scores over a certain value to show what the largest outliers are. This is more useful than ranking by attendance because pool attendance may vary depending on the population size of each pool's area, which can lead to a poor comparison.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would ask what days were used, and when throughout the year were those days pulled from. This is because obviously summer days will have higher attendance than winter days, and weather conditions play a key role in day-to-day pool attendance. A storm in one area may lead to unreliable results for that area's pool attendance that makes the data hard to compare. I would want to assure that the days being sampled are as similar as possible in terms of conditions and point in the year to get a true assessment as to what pools are really underperforming.

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
