---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: Ethan Putnam
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

Based on this scatterplot, a higher percentage of FARMS-eligible students leads to a lower number of students exceeding math expectations. Compared to the scatterplot from the class activity, the data shows a more apparent trend. The data is however still less linear than the data we looked at for the scatterplot we made.


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

The p-value (< 0.0000) tells us that there are statistically significant results here. The Adjusted R-squared that the variation in math scores is 67% explained by FARMS percentage. The coefficient of -0.595 tells us that there is a noticeable negative relationship between these two variables.

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

This tells us that there is a real and significant correlation between higher FARMS percentage and declining math scores. This model is still worth exploring further because while we have determined that there is a significant correlation, we should look to identify the extent of the correlation.

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

The two schools with the largest absolute values of residuals were Fox Chapel Elementary and Broad Acres Elementary. Fox Chapel was nearly 30 over, and Broad Acres was over by over 28.

The zip code of Fox Chapel is 20876. The area has an average income per capita of about three quarters of the average income of the wider DMV region. The poverty rate of 9.8% is 20 percent above the DMV area, and the under 18 poverty rate of 21% is more than double that of the DMV and nearly double that of the state. 

The zip code of Broad Acres (now JoAnn Leleck Elementary School at Broad Acres) is 20904. The area has slightly higher average income per capita to the other zip code, but has a lower poverty rate (9.3%), and a much lower under 18 poverty rate of 13%.

According to Census Reporter, these figures have a margin of error of 10%.

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

Looking at this information, it is clear that more FARMS-eligible students has had at least some correlation with lower math scores. However, it is also true that the highest residuals when looking at math scores were above expected for schools with over 50% FARMS-eligibility. When doing a story about the school board member's statement, it would be important to describe the overall trend that the board member identified, but to also provide context that gives readers a better understanding as to what may be the root cause of these results and why the declarative statement should not be viewed as an absolute truth.
