---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates = md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 5000, property_rate_2019 = property2019/pop2019 * 5000, violent_rate_2020 = violent2020/pop2020 * 5000, property_rate_2020 = property2020/pop2020 * 5000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

In 2019, Baltimore City had the highest rate of violent crime, with over 91 violent crimes committed per 5,000 people, followed by Ocean City (65 per 5,000). Ocean City had the highest rate of property crime in 2019: 518 per 5,000 people, followed by Oakland (366 per 5,000). In 2020, Ocean City had the highest violent crime rate with 119 per 5,000 people, a major increase from 2019's 65 per 5,000. Baltimore City had the second highest violent crime rate with over 79 per 5,000, which marked a decrease from 2019. Ocean City also had the highest rate in property crime in 2020, with a rate of 606 per 5,000 committed. The next highest rate of property crime was committed in Colmar Manor in PG County: about 287 per 5,000 people. The story here is the major increase in crime in Ocean City in 2020, both violent and property.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change = md_cities_rates |> mutate(violent_crime_pct = (((violent2020 - violent2019) / violent2019) * 100), property_crime_pct = (((property2020 - property2019) / property2019) * 100))

```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The overall trend is a decrease in violent crime, despite the most notable outlier in Ocean City, which saw an 82.4% increase in violent crime between 2019 and 2020. Baltimore City, which held the highest violent crime rate in 2019, had a 15.3% decrease in violent crime in 2020. Property crime also saw a general decrease, outside of Prince George's County which had a 105% increase between 2019 and 2020. Ocean City, which had the highest property crime rate in 2020, had a 16.6% increase from 2019. The story in the data is a general decrease in crime, likely as a result of the pandemic, but with some very notable outliers. Ocean City's increased crime in 2020 continues to be a story to watch, as well as the 105% increase in property crime in Prince George's County in 2020.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Between 2019 and 2020, Ocean City saw a rise in both property and violent crime, despite a general decrease in crime across the state.

Throughout much of Maryland, crime decreased in 2020. Baltimore City had Maryland's leading violent crime rate in 2019 with over 91 violent crimes committed per 5,000 people. That number dropped to over 78 violent crimes in 2020, when the city had a -15.3% violent crime dropoff. Ocean City had an 82.4% violent crime increase and 16.6% property crime increase in that time. In 2019, Ocean City had a violent crime rate of over 65 per 5,000 people, and had a property crime rate of 518 per 5,000 people, the highest in the state. In 2020, Ocean City's violent crime rate rose to 119 per 5,000 people. The property crime rate rose to 606 per 5,000 people.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
