---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mmdata <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset

```{r}
set.seed(50)

sample_5 <- mmdata |> sample_n(5)
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the red or black number of M&Ms

```{r}

sample_5 |> summarize(red_mean = mean(red), red_sd = sd(red))

```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
mmdata |> summarize(redmean = mean(red), redsd = sd(red), blackmean = mean(black), blacksd = sd(black))
```


### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

Both the means and standard deviations of red and black M&Ms are very similar. The red and black means are separated by less than 1, as are the red and black standard deviations.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

This data shows that random sampling gets more accurate with a larger sample size. The closer the sample size gets to the size of the true population, the more accurate the results are to the population. In this case, that means getting closer to the true number of M&Ms of each color.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
