---
title: "final_project"
name: Ethan Putnam
---

```{r}
library(tidyverse)
```

# Loading data, updating dataframes

```{r}
balco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Baltimore%20county%20election%20results%20-%20Sheet1.csv")
garrco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Garrett%20county%20election%20results%20-%20Sheet1.csv")
howco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Howard%20county%20election%20results%20-%20Sheet1.csv")
wicco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Wicomico%20county%20election%20results%20-%20Sheet1.csv")

balco_election <- balco_election |> mutate(county = "Baltimore")
garrco_election <- garrco_election |> mutate(county = "Garrett")
howco_election <- howco_election |> mutate(county = "Howard")
wicco_election <- wicco_election |> mutate(county = "Wicomico")

fredco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcoelection.csv")

moco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocoelection.csv")

pgco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/pgelection.csv")

wasco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/washingtonelection.csv")

fredco_election <- fredco_election |> mutate(county = "Frederick")
moco_election <- moco_election |> mutate(county = "Montgomery")
pgco_election <- pgco_election |> mutate(county = "Prince George's")
wasco_election <- wasco_election |> mutate(county = "Washington")

combined_election <- bind_rows(fredco_election, moco_election, pgco_election, wasco_election, balco_election, garrco_election, howco_election, wicco_election)



fredco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcopopulation.csv")

moco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocopopulation.csv")

fredco_pop <- fredco_pop |> mutate(county = "Frederick")
moco_pop <- moco_pop |> mutate(county = "Montgomery")


combined_pop <- bind_rows(fredco_pop, moco_pop)

combined_election <- combined_election |> mutate(dem_margin = REP - DEM)
```


# Combining election and population dataframes

```{r}
election_pop_data <- left_join(combined_election, combined_pop, join_by(county, YEAR==Year))
```

#GGPLOT

```{r}
ggplot(fredco_election, aes(x = year)) +
  geom_line((y = DEM), color = "blue") +
  geom_line((y = REP), color = "red") +
  geom_point() +
  labs(title = "Percentage of Presidential Election Vote Won By Democratic and Republican Candidates in Frederick County (1960-2024)",
       x = "Year",
       y = "Percentage of Vote Won",
       caption = "Source: U.S. Election Atlas")
```


# PREVIOUS WORK BELOW


# Potential tests

Some tests that I think could help to determine if Frederick has become a purple county would be:

- Hypothesis test to determine if Frederick has actually experienced a change in voting results (an increase in Democratic support and a decrease in Republican support). However, I am unsure of how to actually put the code together. 

- A confidence interval test could also be used to determine how much of a change has occurred, if any. I also am not sure how to put this together, or if it would apply as much as a hypothesis test, but I put both just in case.

- GGPlot would be a great tool to use, too. It will be essential to create a graph showing the compared success of Democratic and Republican candidates in Frederick, along with population growth.

- I also want to test for correlation between population increase and the increased success of Democratic candidates, to see how related those two variables are.

I am not fully confident on most of these, and may have missed some that would apply. I have the data loaded in and combined correctly now, and if you have any other suggestions for what to do with this data, please let me know!

# Loading election data uploaded to GitHub

```{r}
fredco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcoelection.csv")

moco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocoelection.csv")

pgco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/pgelection.csv")

wasco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/washingtonelection.csv")
```

This data shows the outcome of all presidential election votes since 1960 in Frederick, Montgomery, Prince George's, and Washington County. 

For each election, I have included the percentage of the vote won by the Democratic and Republican nominee, because I could not find the raw vote counts.

# Citations

This data was pulled in its entirety from the U.S. Election Atlas (https://uselectionatlas.org/). As I said, I would have preferred the raw voting numbers, but I could not find them going back as far as I would like. This is a good source and provides strong data to help with this project nonetheless.

# Loading population data uploaded to GitHub

```{r}
fredco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcopopulation.csv")

moco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocopopulation.csv")
```

This data shows the changes in population in Frederick and Montgomery County from 1960-2025. I did not create datasets for Prince George's or Washington County because those counties have not experienced many changes politically, and thus population change seems irrelevant to the question at hand. 

While Montgomery County has not experienced political shifts like Frederick has, I still pulled the population data to give a contrast to Frederick and to maybe look into later to address whether or not people are leaving Montgomery County due to high cost of living and moving to Frederick, like some people in Frederick have told me.

# Citations

Frederick County data: https://docs.google.com/document/d/154GlaqQJFAn337gbp0lxWuQihw4z66KrhH8LPLNeVxw/edit?usp=sharing

Montgomery County data: https://docs.google.com/document/d/1WYr1rr3D2ipttLgrcydP5x3jWVjamItI-sd9jbLHFRU/edit?usp=sharing

These links navigate to google docs compiling the sources used to assemble the population data.

# Next Steps

After figuring out how to get the data to import correctly, I think that the next step will be to create a visualization to show what political shifts have actually occurred in each county, and then to put Frederick's political shifts side-by-side with population shifts. I will do the same with Montgomery County.

If there is anything else that I should be doing, please let me know!

# Combining files

```{r}
fredco_election <- fredco_election |> mutate(county = "Frederick")
moco_election <- moco_election |> mutate(county = "Montgomery")
pgco_election <- pgco_election |> mutate(county = "Prince George's")
wasco_election <- wasco_election |> mutate(county = "Washington")

combined_election <- bind_rows(fredco_election, moco_election, pgco_election, wasco_election)


fredco_pop <- fredco_pop |> mutate(county = "Frederick")
moco_pop <- moco_pop |> mutate(county = "Montgomery")

combined_pop <- bind_rows(fredco_pop, moco_pop)
```





