---
title: "final_project"
name: Ethan Putnam
---

```{r}
library(tidyverse)
```

# Loading data, updating dataframes

```{r}
balco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Baltimore%20county%20election%20results%20-%20Sheet1.csv")
garrco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Garrett%20county%20election%20results%20-%20Sheet1.csv")
howco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Howard%20county%20election%20results%20-%20Sheet1.csv")
wicco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/Wicomico%20county%20election%20results%20-%20Sheet1.csv")

balco_election <- balco_election |> mutate(county = "Baltimore")
garrco_election <- garrco_election |> mutate(county = "Garrett")
howco_election <- howco_election |> mutate(county = "Howard")
wicco_election <- wicco_election |> mutate(county = "Wicomico")

fredco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcoelection.csv")

moco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocoelection.csv")

pgco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/pgelection.csv")

wasco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/washingtonelection.csv")

fredco_election <- fredco_election |> mutate(county = "Frederick")

fredco_election <- fredco_election |> mutate(dem_margin = DEM - REP)

moco_election <- moco_election |> mutate(county = "Montgomery")
pgco_election <- pgco_election |> mutate(county = "Prince George's")
wasco_election <- wasco_election |> mutate(county = "Washington")

combined_election <- bind_rows(fredco_election, moco_election, pgco_election, wasco_election, balco_election, garrco_election, howco_election, wicco_election)



fredco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcopopulation.csv")

moco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocopopulation.csv")

fredco_pop <- fredco_pop |> mutate(county = "Frederick")
moco_pop <- moco_pop |> mutate(county = "Montgomery")


combined_pop <- bind_rows(fredco_pop, moco_pop)

combined_election <- combined_election |> mutate(dem_margin = DEM - REP)
```


# Combining election and population dataframes

```{r}
election_pop_data <- left_join(combined_election, combined_pop, join_by(county, YEAR==Year))
```

# GGPLOT

```{r}
fredco_election |>
ggplot() +
  geom_line(aes(x = YEAR, y = DEM), color = "blue") +
  geom_line(aes(x = YEAR, y = REP), color = "red") +
  labs(title = "Percentage of Presidential Election Vote Won By Democratic and Republican Candidates in Frederick County (1960-2024)",
       x = "Year",
       y = "Percentage of Vote Won",
       caption = "Source: U.S. Election Atlas")
```

There appears to be an increase in support for Democratic candidates and a decline in support for Republican candidates, but that still needs to be tested further.

# Linear regression (Dem Margin x Year)

```{r}
model <- lm(dem_margin ~ YEAR, data = fredco_election)
summary(model)
```

When examining Frederick County presdidential election data, it is clear from the p-value of 0.1504 that the change in the Democratic voting margin from 1960 to 2024 is statistically insignificant, as it is greater than the 0.05 threshold for statistical significance.

The Adjusted R-squared value of 0.07499 means that about 7.5% of the change in the Democratic voting margin is explained by time, but that was not really the focus of this test. The coefficient of 0.3027 tells us that as time has gone on, the Democratic vote margin in presidential elections in Frederick County has grown.

# Linear regression (Dem Margin x Population)

```{r}

fredco_pop_election <- left_join(fredco_election, fredco_pop, join_by(county, YEAR ==Year))

model <- lm(dem_margin ~ Population, data = fredco_pop_election)
summary(model)
```

When testing the correlation between population and the Democratic vote margin in Frederick County, the p-value of 0.2415 again tells us that these results are not statistically significant. The Adjusted R-squared tells us that about 36.3% of the change in the Democratic vote margin is explained by population, and the coefficient of 1.220e-04 tells us that there is a very slight increase in Democratic vote margin when the population grows.

# Testing Dem. voting margin in other counties

```{r}
moco_election <- moco_election |> mutate(dem_margin = DEM - REP)
moco_pop_election <- left_join(moco_election, moco_pop, join_by(county, YEAR ==Year))

model <- lm(dem_margin ~ Population, data = fredco_pop_election)
summary(model)
```

This test, now looking at Montgomery County, produced similar results. While the high p-value signifies that there has not necessarily been significant change, the positive (though small) coefficient tells us that there has been growth in the Democratic vote margin over time, and the Adjusted R-squared tells us that 36.3% of that change has been the result of population growth.


```{r}
pgco_election <- pgco_election |> mutate(dem_margin = DEM - REP)
model <- lm(dem_margin ~ YEAR, data = pgco_election)
summary(model)

wasco_election <- wasco_election |> mutate(dem_margin = DEM - REP)
model <- lm(dem_margin ~ YEAR, data = wasco_election)
summary(model)
```


The test on Prince George's County prompted the first statistically significant results thus far, with a p-value of 1.1e-06. The coefficient (1.439) tells us that the Democratic vote margin has increased as time has gone on, and the Advanced R-squared tells us that 79.08% of this change is explained by time.

```{r}
balco_election <- balco_election |> mutate(dem_margin = DEM - REP)
model <- lm(dem_margin ~ YEAR, data = balco_election)
summary(model)

garrco_election <- garrco_election |> mutate(dem_margin = DEM - REP)
model <- lm(dem_margin ~ YEAR, data = garrco_election)
summary(model)

howco_election <- howco_election |> mutate(dem_margin = DEM - REP)
model <- lm(dem_margin ~ YEAR, data = howco_election)
summary(model)

wicco_election <- wicco_election |> mutate(dem_margin = DEM - REP)
model <- lm(dem_margin ~ YEAR, data = wicco_election)
summary(model)
```

Here, we found a statistically significant increase in the Democratic vote margin in Baltimore County and Howard County, as well as a statistically significant decrease in Garrett County.

# Notes

While the Frederick County results are not significantly significant according to their p-values, the test is analyzing how much the Democratic vote margin has increased. This does not necessarily equate to if the county has become "purple" which would by nature not have a vast Democratic swing. A purple county is highly contested, so I do not believe these results completely write off Frederick having become purple. Both tests of Democratic vote margin's correlation with time and population showed a positive slope, so there has been some increase. 

I am not sure how I should utilize the main dataframe (election_pop_data) that combines all counties and their respective election and (if applicable) population data. For these tests, I have found working with them individually to make the most sense, and do not know how I should go about testing that dataframe, or what to look for.

# PREVIOUS WORK BELOW




# Potential tests

Some tests that I think could help to determine if Frederick has become a purple county would be:

- Hypothesis test to determine if Frederick has actually experienced a change in voting results (an increase in Democratic support and a decrease in Republican support). However, I am unsure of how to actually put the code together. 

- A confidence interval test could also be used to determine how much of a change has occurred, if any. I also am not sure how to put this together, or if it would apply as much as a hypothesis test, but I put both just in case.

- GGPlot would be a great tool to use, too. It will be essential to create a graph showing the compared success of Democratic and Republican candidates in Frederick, along with population growth.

- I also want to test for correlation between population increase and the increased success of Democratic candidates, to see how related those two variables are.

I am not fully confident on most of these, and may have missed some that would apply. I have the data loaded in and combined correctly now, and if you have any other suggestions for what to do with this data, please let me know!

# Loading election data uploaded to GitHub

```{r}
fredco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcoelection.csv")

moco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocoelection.csv")

pgco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/pgelection.csv")

wasco_election <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/washingtonelection.csv")
```

This data shows the outcome of all presidential election votes since 1960 in Frederick, Montgomery, Prince George's, and Washington County. 

For each election, I have included the percentage of the vote won by the Democratic and Republican nominee, because I could not find the raw vote counts.

# Citations

This data was pulled in its entirety from the U.S. Election Atlas (https://uselectionatlas.org/). As I said, I would have preferred the raw voting numbers, but I could not find them going back as far as I would like. This is a good source and provides strong data to help with this project nonetheless.

# Loading population data uploaded to GitHub

```{r}
fredco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/fredcopopulation.csv")

moco_pop <- read_csv("https://raw.githubusercontent.com/eputnam1/jour405_spring26/refs/heads/main/project/mocopopulation.csv")
```

This data shows the changes in population in Frederick and Montgomery County from 1960-2025. I did not create datasets for Prince George's or Washington County because those counties have not experienced many changes politically, and thus population change seems irrelevant to the question at hand. 

While Montgomery County has not experienced political shifts like Frederick has, I still pulled the population data to give a contrast to Frederick and to maybe look into later to address whether or not people are leaving Montgomery County due to high cost of living and moving to Frederick, like some people in Frederick have told me.

# Citations

Frederick County data: https://docs.google.com/document/d/154GlaqQJFAn337gbp0lxWuQihw4z66KrhH8LPLNeVxw/edit?usp=sharing

Montgomery County data: https://docs.google.com/document/d/1WYr1rr3D2ipttLgrcydP5x3jWVjamItI-sd9jbLHFRU/edit?usp=sharing

These links navigate to google docs compiling the sources used to assemble the population data.

# Next Steps

After figuring out how to get the data to import correctly, I think that the next step will be to create a visualization to show what political shifts have actually occurred in each county, and then to put Frederick's political shifts side-by-side with population shifts. I will do the same with Montgomery County.

If there is anything else that I should be doing, please let me know!

# Combining files

```{r}
fredco_election <- fredco_election |> mutate(county = "Frederick")
moco_election <- moco_election |> mutate(county = "Montgomery")
pgco_election <- pgco_election |> mutate(county = "Prince George's")
wasco_election <- wasco_election |> mutate(county = "Washington")

combined_election <- bind_rows(fredco_election, moco_election, pgco_election, wasco_election)


fredco_pop <- fredco_pop |> mutate(county = "Frederick")
moco_pop <- moco_pop |> mutate(county = "Montgomery")

combined_pop <- bind_rows(fredco_pop, moco_pop)
```





